---
title: 'Causal Inference Demo'
author: "Alex van Vorstenbosch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  beamer_presentation: hanover
  color_theme: dolphin
header-includes:
  - \usepackage{tikz}
  - \usepackage{multicol}
  - \usepackage{hyperref}
  - \hypersetup{
      colorlinks=true,
      linkcolor=blue,
      urlcolor=cyan,
      }
---
<!--

    includes:
      in_header: preamble.tex

--->

```{r setup II, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(eval = TRUE, cache=TRUE, echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(dagitty)
library(brms)
```


# What are we going to see

* Defining the problem
* DAG
* Simulating the data
* Can we use BRMS to retreive the parameters
* \textbf(Real World Data...)
* Is our dag valid?
* Results



# Defining the Problem

* Mathematics is on of THE fundamental school subject...
* ...yet, it is stereotypically seen as boring and difficult.
* In this demo we explore what influences school performance in Math
* ... and possible intervention paths we could use.

# DAG



